Rates and delay times of type Ia supernovae in the Dark Energy Survey

### 2021-03-10 draft.tex

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_DTD_fit_beta_norm_Qerf1.1_split_x1_fitjoint.png}
    \caption{As per Fig. \ref{fig:rate_fitted_split_x1} but with the data fit simultaneously by a single two-component DTD, with fixed slope $\beta = -1$ and a prompt fraction that is 1 at short delay times, 0 at large delay times, and a linear function of time between $t_1$ and $t_2$.%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:rate_fitted_split_x1_2c_DTD}}
\end{figure}
The SN Ia rate as a function of stellar mass for SNe split by stretch is shown in Fig. \ref{fig:rate_fitted_split_x1}. The evolution of the rate with stellar mass is significantly different for the two sub-samples: high-stretch SNe dominate the rate at low stellar mass but tail off in the higher mass galaxies, while low-stretch SNe are subdominant in low mass galaxies but display a much steeper dependence on stellar mass and make up the vast majority of SNe in high-mass hosts. These measurements reflect the observed correlations of $x_1$ with stellar mass and sSFR (and by inference, stellar age) as been seen by \citet{Rigault2013,Graur2017,Rigault2018,Rose2019,Nicolas2020,Rose2021}. The corresponding DTD fit results in a steep power-law $\beta = -1.68 \pm 0.15$ for high-stretch SNe indicative of a population of predominantly prompt SNe, and a much shallower decay slope ($\beta = -0.79 \pm 0.08$) for low-stretch SNe representing a much more delayed population. This result provides and intriguing confirmation of that observed in 101 SDSS SNe Ia by \citet{Brand2010} who found

@@

Alternatively, low- and high-stretch SNe could form via identical evolutionary channels but with a different onset time since the episode of star-formation. The predominantly high-stretch "prompt" SNe would begin exploding as soon as the WDs are formed (nominally 40 Myr) whereas the mainly low-stretch "tardy" SNe would only begin exploding after an extended period of time of the order 1 Gyr. The rates of both populations would follow the fiducial $t^{-1}$ distribution.
Such a scenario may be explained by the sub-$M_{\mathrm{Ch}}$ double-detonation paradigm \citep[e.g.][]{Sim2010,Blondin2017,Shen2017a}. Sub-$M_{\mathrm{Ch}}$ SNe Ia typically involve the merger of two WDs and the SN luminosity can be correlated with the mass of the primary WD; a correlation between primary WD mass and age would then lead to the different rates, and apparent DTDs, of low and high-stretch SNe observed in this work. Such an evolution of light curve properties with age has recently been seen across the full range of "normal" SNe Ia in 1-dimensional simulations by \citet{Shen2021} and is a promising avenue for further investigation although we note that they struggle to reproduce observed light curves at later times ($t\gtrsim30$~d; \citealt{Shen2021,Gronow2021}).

>>>>

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_DTD_fit_beta_norm_Qerf1.1_split_x1_fitjoint.png}
    \caption{As per Fig. \ref{fig:rate_fitted_split_x1} but with the data fit simultaneously by a single two-component DTD, with fixed slope $\beta = -1$ and a prompt fraction that is 1 at short delay times, 0 at large delay times, and a linear function of time between $t_1$ and $t_2$.%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:rate_fitted_split_x1_2c_DTD}}
\end{figure}
The SN Ia rate as a function of stellar mass for SNe split by stretch is shown in Fig. \ref{fig:rate_fitted_split_x1}. The evolution of the rate with stellar mass is significantly different for the two sub-samples: high-stretch SNe dominate the rate at low stellar mass but tail off in the higher mass galaxies, while low-stretch SNe are subdominant in low mass galaxies but display a much steeper dependence on stellar mass and make up the vast majority of SNe in high-mass hosts. These measurements reflect the observed correlations of $x_1$ with stellar mass and sSFR (and by inference, stellar age) as been seen by \citet{Rigault2013,Graur2017,Rigault2018,Rose2019,Nicolas2020,Rose2021}. The corresponding DTD fit results in a steep power-law $\beta = -1.68 \pm 0.15$ for high-stretch SNe indicative of a population of predominantly prompt SNe, and a much shallower decay slope ($\beta = -0.79 \pm 0.08$) for low-stretch SNe representing a much more delayed population. This result provides and intriguing confirmation of that observed in 101 SDSS SNe Ia by \citet{Brandt2010} who found

@@

Alternatively, low- and high-stretch SNe could form via identical evolutionary channels but with a different onset time since the episode of star-formation. The predominantly high-stretch "prompt" SNe would begin exploding as soon as the WDs are formed (nominally 40 Myr) whereas the mainly low-stretch "tardy" SNe would only begin exploding after an extended period of time of the order 1 Gyr. The rates of both populations would follow the fiducial $t^{-1}$ distribution.
Such a scenario may be explained by the sub-$M_{\mathrm{Ch}}$ double-detonation paradigm \citep[e.g.][]{Sim2010,Blondin2017,Shen2017}. Sub-$M_{\mathrm{Ch}}$ SNe Ia typically involve the merger of two WDs and the SN luminosity can be correlated with the mass of the primary WD; a correlation between primary WD mass and age would then lead to the different rates, and apparent DTDs, of low and high-stretch SNe observed in this work. Such an evolution of light curve properties with age has recently been seen across the full range of "normal" SNe Ia in 1-dimensional simulations by \citet{Shen2021} and is a promising avenue for further investigation although we note that they struggle to reproduce observed light curves at later times ($t\gtrsim30$~d; \citealt{Shen2021,Gronow2021}).
